Parallel-in-time Methods for Particle Simulations

### 2025-02-07 paper/main.tex
Line breaks at figure refs

The resulting run times were collected in Table \ref{runtimes}. For visualizations of the simulation results see Figure \ref{sim} and \ref{energy}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal. For more information, see Table \ref{runtimes}.}
\label{sim}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{Conservation of energy in the simulation. Same simulation as in Figure \ref{sim}. With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of time steps.}
\label{energy}
\end{figure}

@@

We can compare this with the real measured speed-up, see Table \ref{runtimes}. We observe that in reality, the program runs around 5 times faster with 16 threads. We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure \ref{speeds}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
\caption{Comparison of the expected run times (red curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red ``x'' point)) and the real measured times with different numbers of CPU cores (blue points). For more information, see Table \ref{runtimes}.}
\label{speeds}
\end{figure}

@@

If we instead approximate the simulation using a similar serial simulation with a lower number of steps, we get a substantially bigger error than the parareal algorithm while not improving the run time significantly. This is also shown in Table \ref{runtimes} (as ``serial (approx.)'').

@@

A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were run and the results are shown in Figure \ref{timetest}. The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.

A key feature observed in Figure \ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes. After every initial downward spike, there is always an upward trend.

@@

\subsection{Speed-up}
The run times of a fixed simulation with different numbers of CPU cores were measured in Experiment I and compared with the theoretical expectations. As was visualized in Figure \ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).

@@

By comparing parareal with an approximating serial run using a bigger step size (see ''serial (approx.)`` in Table \ref{runtimes}), we have also shown that in this case, the parareal algorithm beats simply approximating the solution using a smaller step size in both accuracy and run time.

@@

Finally, in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation. The results are presented in Figure \ref{timetest}. The result is a prediction of the optimal coarse solver step size for our simulation, which is the biggest possible such that the algorithm converges in one iteration.

>>>>

The resulting run times were collected in Table~\ref{runtimes}. For visualizations of the simulation results see Figures~\ref{sim}~and~\ref{energy}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_3.eps}}
\caption{Trajectory computed in the simulation of $10^4\text{ days}$ of the solar system using parareal. For more information, see Table~\ref{runtimes}.}
\label{sim}
\end{figure}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_4.eps}}
\caption{Conservation of energy in the simulation. Same simulation as in Figure~\ref{sim}. With a small enough step size, even a non-symplectic integrator like RK4 can preserve energy for some number of time steps.}
\label{energy}
\end{figure}

@@

We can compare this with the real measured speed-up, see Table~\ref{runtimes}. We observe that in reality, the program runs around 5 times faster with 16 threads. We can also compare the expected speed-up $S_\text{parareal}$ for other numbers of CPUs, see Figure~\ref{speeds}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\linewidth]{fig_5.eps}}
\caption{Comparison of the expected run times (red curve, computed from $S_\text{parareal}$ and the run time of the serial algorithm (red ``x'' point)) and the real measured times with different numbers of CPU cores (blue points). For more information, see Table~\ref{runtimes}.}
\label{speeds}
\end{figure}

@@

If we instead approximate the simulation using a similar serial simulation with a lower number of steps, we get a substantially bigger error than the parareal algorithm while not improving the run time significantly. This is also shown in Table~\ref{runtimes} (as ``serial (approx.)'').

@@

A series of simulations with the same configuration but with different numbers of steps of the coarse solver ($N_\mathcal{C}$) were run and the results are shown in Figure~\ref{timetest}. The parameter for the stopping criterion was fixed at $\varepsilon = 10^{-3}$ and we observe varying numbers of iterations required to reach this criterion.

A key feature observed in Figure~\ref{timetest} are the ``spikes'' in the run times, that occur when a change in the number of required iterations changes. After every initial downward spike, there is always an upward trend.

@@

\subsection{Speed-up}
The run times of a fixed simulation with different numbers of CPU cores were measured in Experiment I and compared with the theoretical expectations. As was visualized in Figure~\ref{speeds}, we can observe that the measured run times show a trend similar to the expected run times (corresponding to the predicted speed-ups).

@@

By comparing parareal with an approximating serial run using a bigger step size (see ''serial (approx.)`` in Table~\ref{runtimes}), we have also shown that in this case, the parareal algorithm beats simply approximating the solution using a smaller step size in both accuracy and run time.

@@

Finally, in Experiment II, we attempted to find the optimal setting (with the smallest run time and a fixed $\varepsilon$) of the course solver step size for our simulation. The results are presented in Figure~\ref{timetest}. The result is a prediction of the optimal coarse solver step size for our simulation, which is the biggest possible such that the algorithm converges in one iteration.
